A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-07-19 doc/jcomp2012/jcomp2012.tex
updated jcomp abstract
updates to section 3

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation in 3
    dimensions. We show that this method can be an effective solver
    for sparse matrix systems and is competitive with deterministic
    methods including preconditioned Conjugate Gradient. We also
    discuss various aspects of preconditioning and its general
    applicability to broader classes of problems.

@@

The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
error instead of the direct method outlined in
\S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
adjoint method over the direct method, we choose the two-dimensional
time-dependent Poisson equation as a simple model problem:

@@

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in several orders of magnitude
speedup over the direct solver while the number of iterations required
to converge is of a similar scale. We expect this for several
reasons. First, with an equivalent number of histories specified for
both solvers and a system of size $N \times N$, the direct solver will
compute $N \times N$ random walks per permuation while the adjoint
solver will only compute one. Initially, this may not seem like a fair
comparison, however, we see from Fig.~\ref{fig:poisson_iterations}
that the number of iterations requried to converge is approximately
the same. This is necessary in the direct method to ensure a
contribution from each state as the random walk sequence will only
contribute to the starting state. For the adjoint method, because the
random walk sequence contributes to the state in which it currently
resides, fewer histories are neccessary to compute contributions from
all the nonzero parts of the source. Therefore, although the total
number of random walks per iteration is in fact larger for the direct
method, the solution computed at each iteration is no better than that
computed by the adjoint method.

Although CPU time is shown to be greatly decreased by using the
adjoint solver in Fig.~\ref{fig:poisson_cpu_time}, the iteration
behavior was approximately the same. To assess the convergence
properties of MCSA using each solver and stencil, the infinity norm of
the residual computed in Eq.~(\ref{eq:general-MCSA}) was collected at
each iteration for a fixed problem
size. Fig.~\ref{fig:poisson_convergence} gives the results of these
computations. First, it is worthy to note on the semilog plot that we
are indeed achieving the expected exponential convergence from
MCSA. Second, we note that both the direct and adjoint methods have
approximately equivalent convergence behavior, leading to CPU time
being the dominating factor in method selection.

>>>>

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation in 3
    dimensions. We show that this method can be an effective solver
    for sparse matrix systems and is competitive with deterministic
    methods including preconditioned Conjugate Gradient. We also
    discuss various aspects of preconditioning the method and its
    general applicability to broader classes of problems.

@@

The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
error instead of the direct method outlined in
\S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
adjoint method over the direct method within the context of MCSA, we
choose the two-dimensional time-dependent Poisson equation as a simple
model problem:

@@

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in several orders of magnitude
speedup over the direct solver while the number of iterations required
to converge is of a similar scale. We expect this for several
reasons. First, with an equivalent number of histories specified for
both solvers and a system of size $N \times N$, the direct solver will
compute $N \times N$ random walks per permuation while the adjoint
solver will only compute one. This is necessary in the direct method
to ensure a contribution from each state as the random walk sequence
will only contribute to the starting state. For the adjoint method,
because the random walk sequence contributes to the state in which it
currently resides, fewer histories are neccessary to compute
contributions from all the nonzero parts of the source. Initially,
this may not seem like a fair comparison, however, we see from
Fig.~\ref{fig:poisson_iterations} that the number of iterations
requried to converge is approximately the same and therefore the extra
computations performed by the direct method do not improve the error
estimation. Therefore, although the total number of random walks per
iteration is in fact larger for the direct method, the solution
computed at each iteration is no better than that computed by the
adjoint method.

As an additional comparision, the convergence behavior of MCSA can be
analyzed using both solvers to detect any performance benefits. To
assess the convergence properties of MCSA using each solver and
stencil, the infinity norm of the residual computed in
Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a
fixed problem size. Fig.~\ref{fig:poisson_convergence} gives the
results of these computations. First, it is worthy to note on the
semilog plot that we are indeed achieving the expected exponential
convergence from MCSA. Second, we note that both the direct and
adjoint methods have approximately equivalent convergence behavior,
leading to CPU time being the dominating factor in method selection.
